---
title: "BTYD CLV Model implementation"
author: "Nandini Hegde"
date: "August 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Predicting Customer Lifecycle value (CLV) of a customer in a Non-contractual setting using BTYD/BTYDplus models

This is project demontrating the use of BTYD models for predicting future CLV of a customer based on historical transactions and sales value.
Link to the original paper: https://cran.r-project.org/web/packages/BTYD/vignettes/BTYD-walkthrough.pdf

The BTYDplus package has inbuilt functions and models in addition to the BTYD package. Hence, I will be using it to conduct the analysis
Link to the reference paper: https://cran.r-project.org/web/packages/BTYDplus/vignettes/BTYDplus-HowTo.pdf

## Model selection in a Non-contractual setting

Pareto-NBD and BG-NBD Models can be used in situations when the the transaction opportunities are continuous and not distinct. This is similar to a 
e-commerce website or a direct to customer retail shop.However,BG-NBD models assume that a customer does not die dueing period of inactivity which seems counterintuitive especially with respect to repeat transactions.

BG/BB Model on the other hand is used in situations wherein customers have discrete transaction opportunities akin a charity donation or an auction.


I will be first demonstrating the application of Pareto-NBD and BG-NBD models on the first dataset which is a an online direct to customer sales dataset
I will then go on to use a second dataset to demonstrate the BG/BB model

### Data exploration- Retail Dataset
For the implementation of these models we need an event log constiting of customer id, date/time of transaction,  total sales.
The dataset used in this example is borrowed from:https://www.kaggle.com/regivm/retailtransactiondata#Retail_Data_Transactions.csv

```{r}

#Importing required libraries

install.packages("BTYDplus")
install.packages("BTYD")

library("BTYD")
library("BTYDplus")
library('data.table')
library(lubridate)


#read in data and format as per package requirements
retail<-read.csv("Retail_Data_Transactions.csv")

elog<-retail
#rename columns 
names(elog)<-c("cust","date","sales")
elog$date <- as.Date(elog$date, format = "%m/%d/%Y")

#reorder according to date
elog<-elog[order(elog$cust,elog$date),]

# A snapshot of the data
head(elog)

#checking the format 
str(elog)

#Merge transactions for each customer on each date
elog <- dc.MergeTransactionsOnSameDate(elog)

set.seed(123)
# plot timing patterns of 30 sampled customers
plotTimingPatterns(elog, n = 30, T.cal = median(elog$date),
headers = c("Past", "Future"), title = "")


```
Further on in the analysis, we will divide the dataset into holdout and caliberation period based on the median date ("2013-04-14"). This is done to evaluate the accuracy of the model


```{r}
#Calculating Number of repeated transactions by a customer
n_occur <- data.frame(table(elog$cust))
head(n_occur,5)

op <- par(mfrow = c(1, 2), mar = c(2.5, 2.5, 2.5, 2.5))
# incremental weekly transactions total and repetitive
weekly_inc_total <- elog2inc(elog, by = 7, first = TRUE)
weekly_inc_repeat <- elog2inc(elog, by = 7, first = FALSE)
plot(weekly_inc_total, typ = "l", frame = FALSE, main = "Incremental")
lines(weekly_inc_repeat, col = "red")
legend(1, 25, legend=c("total transactions", "repeat transactions"),col=c( "black","red"), lty=1, cex=0.8)

# cumulative  weekly transactions total and repetitive
weekly_cum_total <- elog2cum(elog, by = 7, first = TRUE)
weekly_cum_repeat <- elog2cum(elog, by = 7, first = FALSE)
plot(weekly_cum_total, typ = "l", frame = FALSE, main = "Cumulative")
lines(weekly_cum_repeat, col = "red")
legend(1, 25, legend=c("total transactions", "repeat transactions"),col=c( "black","red"), lty=1, cex=0.8)
par(op)


```
The x-axis represents time measured in weeks, thus we see that the customers were observed
over a five year time period. The gap between the red line (repeat transactions)
and the black line (total transactions) illustrates the customers’ initial transactions.
These only occur within the first 75 weeks, perhaps because this dataset was subsetted to include customers acquired in that period.


### Data preprocessing- Retail Dataset
```{r}

# date range
range(elog$date)

# date median to split holdout and caliberation period
median(elog$date)


#convert to CBS format
retailCBS <- elog2cbs(elog,T.cal = median(elog$date))
#remove outliers from the CBS file if any customer has an unnaturally high number of repeat transactions (x) as it can cause error during parameter estimation, this particular dataset had none
head(retailCBS)

#Estimating regularity
op <- par(mfrow = c(1, 2))
(k.wheat <- estimateRegularity(elog, method = "wheat",
plot = TRUE, title = "Wheat & Morrison"))
#>[1] 1.037254
(k.mle <- estimateRegularity(elog, method = "mle",
plot = TRUE, title = "Maximum Likelihood"))
#>[1] 1.132881
par(op)


```

Some models are capable of leveraging regularity within transaction timings for improving forecast accuracy.
The Wheat and morrison estimater reports a regularity of close to 1 supports the assumption of exponentially distributed
intertransaction times, and the mle calculation of close to 1 also indicates no stronger degrees of regularity for any subset of highly
active customers

### BG-NBD Model- Retail Dataset
```{r}

# estimate BG-NBD parameters
round(params.BGnbd <- BTYD::bgnbd.EstimateParameters(retailCBS), 3)
#>  r     alpha       a        b 
#>40.704  450.277    0.001 1575.336

# report log-likelihood
bgnbd.cbs.LL(params.BGnbd, retailCBS)
#> [1] -187902.7

# calculate expected future transactions for customers who've
# had 1 to 5 transactions in first 52 weeks across 100 weeks
est5.nbd <- bgnbd.ConditionalExpectedTransactions(params.BGnbd,T.star = 52, x = 1:5,t.x = 100,  T.cal = 52)
for (i in 1:5) {
cat("x =", i, ":", sprintf("%5.3f", est5.nbd[i]), "\n")
}

# x = 1 : 4.318 
# x = 2 : 4.421 
# x = 3 : 4.525 
# x = 4 : 4.628 
# x = 5 : 4.732 

# predict whole customer cohort
retailCBS$xstar.BGnbd <- bgnbd.ConditionalExpectedTransactions(params = params.BGnbd, T.star = 100,x = retailCBS$x, T.cal = retailCBS$T.cal, t.x = retailCBS$t.x)
# compare predictions with actuals at aggregated level
rbind(`Actuals` = c(`Holdout` = sum(retailCBS$x.star)),
`NBD` = c(`Holdout` = round(sum(retailCBS$xstar.BGnbd))))

#         Holdout
# Actuals   62045
# NBD       62221


```

### Pareto-NBD Model- Retail Dataset
```{r}
#convert to CBS format
retailCBS <- elog2cbs(elog,T.cal = median(elog$date))

# estimate Pareto/NBD parameters
params.pnbd <- BTYD::pnbd.EstimateParameters(retailCBS)
names(params.pnbd) <- c("r", "alpha", "s", "beta")
round(params.pnbd, 3)
#> r      alpha        s    beta
#> 40.648 449.612   0.000  43.498 
# report log-likelihood
BTYD::pnbd.cbs.LL(params.pnbd, retailCBS)
#> [1] -187902.7

# For one, we can note, that the maximized log-likelihood of Pareto/NBD is bigger than
# for the BG-NBD model, implying that its data fit is better. 

#Plotting the  distribution of the estimated parameters
pnbd.PlotTransactionRateHeterogeneity(params.pnbd)
pnbd.PlotDropoutRateHeterogeneity(params.pnbd)


# calculate expected future transactions for customers who've
# had 1 to 5 transactions in first 12 weeks, but then remained
# inactive for 40 weeks
est5.pnbd <- BTYD::pnbd.ConditionalExpectedTransactions(params.pnbd,T.star = 52, x = 1:5, t.x = 12, T.cal = 52)
for (i in 1:5) {
  cat("x =", i, ":", sprintf("%5.3f", est5.pnbd[i]), "\n")
}

# x = 1 : 4.317 
# x = 2 : 4.421 
# x = 3 : 4.524 
# x = 4 : 4.628 
# x = 5 : 4.732 

# P(alive) for customers who've had 1 to 5 transactions in first
# 12 weeks, but then remained inactive for 40 weeks
palive.pnbd <- BTYD::pnbd.PAlive(params.pnbd,x = 1:5, t.x = 12, T.cal = 52)
for (i in 1:5) {
  cat("x =", i, ":", sprintf("%5.2f %%", 100*palive.pnbd[i]), "\n")
}

# x = 1 : 99.99 % 
# x = 2 : 99.99 % 
# x = 3 : 99.99 % 
# x = 4 : 99.99 % 
# x = 5 : 99.99 % 



# predict whole customer cohort
retailCBS$xstar.pnbd <- BTYD::pnbd.ConditionalExpectedTransactions(  params = params.pnbd, T.star = 100,x = retailCBS$x, t.x = retailCBS$t.x,T.cal = retailCBS$T.cal)
# compare predictions with actuals at aggregated level
rbind(`Actuals` = c(`Holdout` = sum(retailCBS$x.star)),
      `Pareto/NBD` = c(`Holdout` = round(sum(retailCBS$xstar.pnbd))))
#    Holdout
# Actuals      62045
# Pareto/NBD   62226
#gives almost similar estimates compared to BG-NBD


````